---
title       : Word Counter
subtitle    : Project for Developing Data Products
author      : Sur Tan (22 Nov 2015)
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## About Word Counter
----
'Words Counter' is a simple shiny application to count the number of words and characters of a text. It also finds the top 10 words most frequently used and the percentage each word is used within the text.

https://sur-tan.shinyapps.io/shinyapp

How to Use:
- Type or paste the text in the input text box.
- Click 'Count' button. 
- The output will be displayed in the 'Summary' section.

Limitations:

'Word Counter' makes its own assumption to define the meaning of a word. Refer to next slide on those assumptions.  

---

## How It Works
----
'Word Counter' makes the following assumptions to process the text: 

- Space is the separator of words. 
- Carriage return separates two different paragraphs, so it is converted to a space to separate the last word from the current paragraph to the first word of the next paragrah.
- Double, leading or trailing spaces are removed first as they will include as an additional count when there is no word tie to these spaces.

- Carriage return is not considered as a character, so it is removed before counting number of characters.
- Some words are together with other special characters such as question mark, apostrophe, etc, e.g. 'How?'. When counting the top 10 frequently used words, these will be removed so that 'How' and 'How?' are considered the same words. Note: only a few special characters are included here: ? , . ! " ' ( )  
- The words are all changed to small case before counting.

---

## Codes Example
----
```{r echo=FALSE, warning=FALSE}
library(shiny)
library(plyr)

## Word Counter Server
##
## This is the server processing of the shiny application to count the number of words and characters of a text.
## It will take in the input text and count the number of words, number of characters and find the top 10 of the
## most frequently used words and return the results in the output fields. 
##
## It consists of four functions to process the inputs: 
## (1) extractwords 
## (2) countWords 
## (3) countChars
## (4) findtop10

## extractwords
## This function extracts words from the input text. 
##
## It uses a space to separate each word. Multiple spaces and carriage return will be ignored. 
## It also remove the special characters such as period, comma, single quote, double quote, brackets, 
## question mark and exclation marks from the word. Note: The list of special characters is not 
## exhausted and can be improved. 
## 
## parameter: text - text to process
## return: a vector of the words extracted from the input text
#
extractWords <- function(text) {
    ## remove end of line and replace by space
    text <- gsub("[\r\n]", ' ', text)

    ## remove double spaces
    text <- gsub(' {2,}', ' ', text)
    
    ## remove special characters
    text <- gsub('[(]','',text)
    text <- gsub('[)]','',text)
    text <- gsub('[.]','',text)
    text <- gsub('[,]','',text)
    text <- gsub('[!]','',text)
    text <- gsub('[?]','',text)
    text <- gsub('[\']','',text)
    text <- gsub('[\"]','',text)
    text <- tolower(text)
    
    ## remove leading and trailing spaces
    text <- gsub("[ ]+$", "", text)
    text <- gsub("^[ ]+", "", text)
    
    ## split the text to each word, separated by spaces
    strsplit(text, ' ') 
}

## countWords
## This function counts the number of words from the input text. 
## 
## It simply uses extractwords function to extract words from the input text and do the count.
##
## parameter: text - text to process
## return: number of words in the input text
##
countWords <- function(text) {
    ## extract each word from the text
    words <- extractWords(text)
    
    ## get number of the words
    length(words[[1]])
}

## countChars
## This function counts the number of characters from the input text. 
## 
## It simply removes the carriange return from the text before finding the number of characters.
##
## parameter: text - text to process
## return: number of characters in the input text
##
countChars <- function(text) {
    ## remove carriage returns before count number of characters
    text <- gsub("[\r\n]", '', text)
    
    ## get the number of characers
    nchar(text)
}

## findtop10
## This function finds the top 10 frequently used words from the input text.
## 
## It will return the text "Not Avaialble" if input text is blank. It uses extractwords function to extract 
## words from input text first, then do the count and find the percentage. it will sort the result by the 
## count in descending order and return the first 10 rows.
##
## parameter: text - text to process
## return: a data frame of the top 10 words. The data frame contains 3 columns, Word, Count and Percentage.
##
findtop10 <- function(text) {
  if (text == "") 
     "Not Available"
  else {
    ## convert words to data frame
    words <- extractWords(text)
    df <- data.frame(words)
    
    # count each word occurance 
    counttable <- table(words)
    df <- data.frame(counttable)
    names(df) <- c("word", "count")
    
    ## create exta column to calculate the percentage of the occurences
    df <- cbind(df, paste(round((df$count/sum(df$count))*100,2),'%'))
    
    ## rename the columns
    colnames(df) <- c("Word", "Count", "Percent")
    
    ## order the data by count in descending order
    df <- arrange(df, desc(Count))
    
    ## get the top 10
    head(df, n=10)
  }
}
````

The codes are too long to be shown here. Refer to ui.R and server.R at https://github.com/sur-tan/DevelopingDataProduct for the complete source codes. 

This is the example how the three main functions (countWords, countChars and findtop10) are called with the sample input text to produce the results.

```{r} 
sample_text <- "This is my project for the Developing Data Proudct course, 
Coursera Data Science Specialization. The project is due on 22 Nov 2015. 
It is a simple shiny application that counts the number of words, characters
and find the top 10 frequently used words and its percentage."

numChars <- countChars(sample_text)
numWords <- countWords(sample_text)

````

The number of characters: ***`r numChars `***.
The number of words: ***`r numWords `***.


---

## Codes Example
----

The top 10 frequently used words are: 
```{r}
findtop10(sample_text)
````
